Bayesian Rate Estimation

### 2012-03-31 many.tex
Updates to many.tex.  Added section describing analytic model and recovery of rates.

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}

@@

@@

We assume that we are presented with a data set of $N$ events.  Each
event may be due to either a signal of interest or an uninteresting
background.  Each event is associated with a ranking statistic, $x$.
Our data set therefore consists of the ranking statistics for the set
of events:
\begin{equation}
  d = \{ x_i | i = 1, \ldots, N \}.
\end{equation}
We assume that the events are sorted by ranking statistic, so that $i
< j$ implies that $x_i < x_j$.

@@

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 or 1, indicating a background or foreground event,
respectively.  These ``state'' flags are additional parameters in our
model.  We can marginalize over our uncertainty in the state of any
given event by summing all posteriors over $f_i = \{0,1\}$.  Given an
identification of each event as foreground or background, and the
rates of each, the likelihood%
\footnote{See \S \ref{sec:likelihood-derivation} for a derivation of
  Eq.~\eqref{eq:likelihood}.} %
of our data is
\begin{multline}
  \label{eq:likelihood}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)
  \right) = \\ \left[ \prod_{\left\{ i | f_i = 1 \right\}} f\left( x_i
    \right) \right]  \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    b\left( x_j \right) \right] \\
  \times \exp\left[ - F\left( \infty \right) \right] 
  \exp\left[ -B\left( \infty \right) \right].
\end{multline}
Written in terms of the rate parameters, this becomes
\begin{multline}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)\right) = \\ R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right],
\end{multline}
where $N_f$ is the number of the $f_i$ that are 1 (i.e.\ the number of
assumed foreground events), and $N_b$ is the number of the $f_i$ that
are 0 (i.e.\ the assumed number of background events).  Note that $N_f
+ N_b = N$, as each event is considered either foreground or
background in our model.

@@

\fixme{Discussion of the various priors one might want to impose on
  the rates, $R_f$ and $R_b$, such as a flat prior on rate, the
  Poisson Jeffrey's prior (which is $\propto 1/\sqrt{R}$?), .}

@@

Putting everything together, the posterior probability of each event's
state and the overall rate is given by
\begin{multline}
  p(R_f, R_b, \left\{ f_i \right\} | d) \propto R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right] p(R_f, R_b).
\end{multline}
If we marginalize over all the flags, $f_i$, then we obtain
\begin{multline}
  p(R_f, R_b | d) \propto \prod_{i} \left[ R_f \hat{f}\left( x_i
    \right) + R_b \hat{b}\left( x_i \right) \right] \\ \times
  \exp\left[ - \left( R_f + R_b \right) \right] p\left( R_f, R_b
  \right).
\end{multline}
If, instead, we marginalize over all but one flag, $f_k$, we obtain
that the posterior ratio of signal to noise probability for event $k$
given the rates $R_f$ and $R_b$ is 
\begin{equation}
  \frac{p(R_f, R_b, f_k = 1 | d)}{p(R_f, R_b, f_k = 0 | d)} =
  \frac{R_f \hat{f}\left( x_k \right)}{R_b \hat{b}\left( x_k \right)}. 
\end{equation}

@@

\begin{acknowledgments}
  Richard O'Shaughnessy for discussions.
\end{acknowledgments}

@@

Consider a set of samples, $\{ x_i \}$, from an

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

@@

\newcommand{\xmin}{x_\mathrm{min}}

\DeclareMathOperator{\erf}{erf}

@@

We assume that we are presented with a data set of $N$ events.  Each
event may be due to either a signal of interest or an uninteresting
background.  Each event is associated with a ranking statistic, $x$.
Our data set therefore consists of the ranking statistics for the set
of events:
\begin{equation}
  d = \{ x_i | i = 1, \ldots, N \}.
\end{equation}

@@

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 or 1, indicating a background or foreground event,
respectively.  These ``state'' flags are additional (unobserved) data
in our model.  We can marginalize over our uncertainty in the state of
any given event by summing all posteriors over $f_i = \{0,1\}$.  Given
an identification of each event as foreground or background, and the
rates of each, the likelihood%
\footnote{See \S \ref{sec:likelihood-derivation} for a derivation of
  Eq.~\eqref{eq:likelihood}.} %
of our data is
\begin{multline}
  \label{eq:likelihood}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)
  \right) = \\ \left[ \prod_{\left\{ i | f_i = 1 \right\}} f\left( x_i
    \right) \right]  \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    b\left( x_j \right) \right] \\
  \times \exp\left[ - F\left( \infty \right) \right] 
  \exp\left[ -B\left( \infty \right) \right].
\end{multline}
Written in terms of the rate parameters, this becomes
\begin{multline}
  \label{eq:likelihood-rates}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)\right) = \\ R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right],
\end{multline}
where $N_f$ is the number of the $f_i$ that are 1 (i.e.\ the number of
assumed foreground events), and $N_b$ is the number of the $f_i$ that
are 0 (i.e.\ the assumed number of background events).  Note that $N_f
+ N_b = N$, as each event is considered either foreground or
background in our model.

It will occasionally be convenient to work with the ratio,
$\hat{f}(x)/\hat{b}(x)$ instead of the values of $\hat{f}$ and
$\hat{b}$ separately.  This ratio can be interpreted as a likelihood
ratio between the foreground and background models for an event that
has occurred with ranking statistic $x$:
\begin{equation}
  \label{eq:loudness-likelihood}
  \frac{\hat{f}(x)}{\hat{b}(x)} = \frac{p(x|\mathrm{foreground})}{p(x|\mathrm{background})}.
\end{equation}
In some circumstances, it is possible to compute this ratio
empirically (for example, in \citet{Cannon2012}, the authors give a
practical method for estimating the distribution of loudnesses for
known-background events and known-foreground events in a mock data
stream from the Advanced LIGO and Advanced Virgo gravitational wave
detectors).  Dividing Eq.~\eqref{eq:likelihood-rates} by
$\prod_{\left\{ i | f_i = 0 \right\}} \hat{b}(x_i)$, we obtain
\begin{multline}
  \label{eq:likelihood-ratio}
  p\left(d, \left\{ f_i \right\} | f(x), b(x) \right) \propto  \\ \left[
    \prod_{\left\{ i | f_i = 1 \right\}}
    \frac{\hat{f}(x_i)}{\hat{b}(x_i)} \right] R_f^{N_f} R_b^{N_b}
  \exp\left[ - \left( R_f + R_b \right) \right].
\end{multline}

@@

Because the overall rates enter the likelihood in
Eq.~\eqref{eq:likelihood-rates} in the same form as in the
constant-rate Poisson likelihood, a reasonable prior for $R_f$ and
$R_b$ is the Poisson Jeffrey's prior%
\footnote{Recall that the Jeffrey's prior for a parameter $\theta$
  with likeilhood function, $p(d|\theta)$, is
  \begin{equation}
    p(\theta) \propto \sqrt{\left\langle \left( \frac{\partial
          p(d | \theta) }{\partial \theta} \right)^2 \right\rangle},
  \end{equation}
  where the expectation is taken over all data, $d$.  The Jeffrey's
  prior assigns prior weight to locations in parameter space
  proportional to the Fisher information that is expected to result
  from a measurement at that parameter value.}, %
\begin{equation}
  \label{eq:Jeffrey-prior}
  p(R) \propto \frac{1}{\sqrt{R}}.
\end{equation}
One advantage of the Jeffrey's prior in this circumstance is that it
has finite probability mass as $R \to 0$, and the likelihood in
Eq.~\eqref{eq:likelihood-rates} controls its behavior as $R\to \infty$
to ensure a normalizable posterior.

Of course, other priors could be used; in particular, previous
experiments may have placed constraints on the rate that can be used
as priors in subsequent experiments.  Throughout the remainder of this
paper, however, we use the Jeffrey's prior,
Eq.~\eqref{eq:Jeffrey-prior}.

@@

Putting everything together, the posterior probability of each event's
state and the overall rate is given by
\begin{multline}
  p(R_f, R_b, \left\{ f_i \right\} | d) \propto R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right] p(R_f, R_b).
\end{multline}
If we marginalize over all the flags, $f_i$, then we obtain
\begin{multline}
  \label{eq:rates-marginalized}
  p(R_f, R_b | d) \propto \prod_{i} \left[ R_f \hat{f}\left( x_i
    \right) + R_b \hat{b}\left( x_i \right) \right] \\ \times
  \exp\left[ - \left( R_f + R_b \right) \right] p\left( R_f, R_b
  \right).
\end{multline}
If, instead, we marginalize over all but one flag, $f_k$, we obtain
that the posterior ratio of signal to noise probability for event $k$
given the rates $R_f$ and $R_b$ is 
\begin{equation}
  \label{eq:foreground-marginalized}
  \frac{p(R_f, R_b, f_k = 1 | d)}{p(R_f, R_b, f_k = 0 | d)} =
  \frac{R_f \hat{f}\left( x_k \right)}{R_b \hat{b}\left( x_k
    \right)}p\left(R_f, R_b\right). 
\end{equation}
Note that the \emph{posterior} estimate of the ratio of signal to
noise probabilities for event $i$ differs from the likelihood ratio,
Eq.~\eqref{eq:loudness-likelihood}, because it depends on the overall
rates of foreground and background events.

@@

\subsection{Analytic Example}
\label{sec:analytic-GW-example}

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering against a set of $N$ templates.  The data
stream consists of stationary Gaussian noise with a power spectral
density $S(f)$ combined additively with some number of signals.  The
signal to noise ratio (SNR) of a template, $h(f)$, given data, $d(f)$,
is given by
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
        \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that 
\begin{equation}
  \left\langle h_1, h_2 \right\rangle \sim 0.
\end{equation}

For a data stream of pure noise, $d(f) = n(f)$, the SNR of a each
template follows a $N(0,1)$ distribution.  If our ranking statistic
for events in each interval of data is their maximum SNR over the $N$
templates, 
\begin{equation}
  x \equiv \max_h \rho_h,
\end{equation}
then the background distribution is
\begin{equation}
  \hat{B}(x) = \left( \frac{1 + \erf\left( \frac{x}{2}
      \right)}{2} \right)^N
\end{equation}
If, in addition, we impose a threshold ranking statistic, $\xmin$, to
be considered an ``event,'' then $\hat{B}$ becomes
\begin{equation}
  \label{eq:analytic-background-rate}
  \hat{B}(x) = \frac{\left( 1 + \erf\left( \frac{x}{2} \right)
    \right)^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N}{2^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N }
\end{equation}

The SNR of a gravitational wave signal in an interferometric detector
scales as $1/d$, where $d$ is the distance to the source.  Ignoring
possible cosmological effects, the number of sources will scale as
$d^3$.  Thus, we expect that the foreground cumulative distribution of
events will follow
\begin{equation}
  \label{eq:analytic-foreground-rate}
  \hat{F}(x) = 1 - \frac{\xmin^3}{x^3}.
\end{equation}

To demonstrate the effectiveness of our formalism, we applied it to a
synthetic data set with foreground and background distributions drawn
from Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} with $R_f^\mathrm{true} = 10.4$
and $R_b^\mathrm{true} = 95.1$.  The cumulative distribution for the
ranking statistic of the synthetic data we used appears in Figure
\ref{fig:analytic-data-cumulative}.

\begin{figure}
  \includegraphics[width=\columnwidth]{data}
  \caption{\label{fig:analytic-data-cumulative} The cumulative
    distribution of the ranking statistics for the synthetic data used
    to test the the formalism on the model from \S
    \ref{sec:analytic-GW-example}.  The solid line gives the
    cumulative distribution of the synthetic data; the dashed line
    gives the theoretical cumulative distribution for the models in
    Eqs.~\eqref{eq:analytic-background-rate} and
    \eqref{eq:analytic-foreground-rate} combined with $R_f = 10.4$ and
    $R_b = 95.1$.}
\end{figure}

In Figure \ref{fig:analytic-rate-recovery}, we show the marginalized
posterior densities for the foreground and background rates recovered
from a synthetic data set drawn from the distributions in
Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} with .  (Refer to
Eq.~\eqref{eq:rates-marginalized}.)  Figure
\ref{fig:analytic-rate-foreground-probs} shows the posterior
foreground probability for each event marginalized over all other
events' types and the foreground and background rates.  (Refer to
Eq.~\eqref{eq:foreground-marginalized}.)

\begin{figure}
  \includegraphics[width=\columnwidth]{rates}
  \caption{\label{fig:analytic-rate-recovery} The posterior densities
    for $R_f$ (in blue) and $R_b$ (in black) for the analytic model
    discussed in \S \ref{sec:analytic-GW-example}.  The vertical lines
    indicate the ``true'' used to generate the synthetic data set.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters.  True foreground events are in blue, background events
    in red.  The solid line is the likelihood ratio
    $p(x|\mathrm{foreground})/p(x|\mathrm{background})$ (see
    Eq.~\eqref{eq:loudness-likelihood}) for this model; this exceeds
    the marginalized foreground probability for many of the events
    because in this data set there are approximately nine times as
    many background events as foreground events (see
    Eq.~\eqref{eq:foreground-marginalized}). }
\end{figure}

\begin{acknowledgments}
  Kipp Cannon, Chad Hanna, and Drew Keppel for providing the data set
  from \citet{Cannon2012} and assisting us in its use. Richard
  O'Shaughnessy for discussions.
\end{acknowledgments}

@@

\fixme{To do.}
